---
title: "Airline Delay and Cancellation"
author: "Group 5"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F,results='markup')

# importing libraries
library(ezids)
library(ggplot2)
library(tidyverse)
library(dplyr)
```

Importing data
```{r}
airlines <- data.frame(read.csv("data/airlines.csv"))
airports <- data.frame(read.csv("data/airports.csv"))
#flights <- data.frame(read.csv("C:/Users/Aditya Kumar/OneDrive/Study/GWU - Data Science/DATS 6101 Intro to Data Science/Project 02/archive (1)/2015.csv"))
#nrow(flights)
```

Our entire flights data for 2015 has 5,819,079 rows which is extremely large. For the sake of our analysis, we will only randomly select 581,908 rows which is 10% of the original dataset size.    
Sub-setting dataset because it is very large. 
```{r}
#set.seed(333)
#sample_size = round(nrow(flights)*0.1)
#flights_sample <- flights[sample(nrow(flights), sample_size, replace = F),]
#head(flights_sample)
```

Exporting smaller sample
```{r}
#write.csv(flights_sample, file = "data/flights_sample.csv",row.names=F)

flight <- read.csv("data/flights_sample.csv")
```

```{r}
xkabledplyhead(flight,title="The first 5 rows of the dataset")
str(flight)
```

This data now has `r nrow(flight)` rows.   

Now we'll begin our analysis.   

# Pre-processing and initial analysis   

Glossary of all the columns in the dataset:   

* FL_DATE = Date of the Flight
* OP_CARRIER = Airline Identifier
* OP_CARRIER_FL_NUM = Flight Number
* ORIGIN = Starting Airport Code
* DEST = Destination Airport Code
* CRS_DEP_TIME = Planned Departure Time
* DEP_TIME = Actual Departure Time
* DEP_DELAY = Total Delay on Departure in minutes
* TAXI_OUT = The time duration elapsed between departure from the origin airport gate and wheels off
* WHEELS_OFF = The time point that the aircraft's wheels leave the ground
* WHEELS_ON = The time point that the aircraft's wheels touch on the ground
* TAXI_IN = The time duration elapsed between wheels-on and gate arrival at the destination airport
* CRS_ARR_TIME = Planned arrival time
* ARR_TIME = Actual Arrival Time = ARRIVAL_TIME - SCHEDULED_ARRIVAL
* ARR_DELAY = Total Delay on Arrival in minutes
* CANCELLED = Flight Cancelled (1 = cancelled)
* CANCELLATION_CODE = Reason for Cancellation of flight: A - Airline/Carrier; B - Weather; C - National Air System; D - Security
* DIVERTED = Aircraft landed on different airport that the one scheduled
* CRS_ELAPSED_TIME = Planned time amount needed for the flight trip
* ACTUAL_ELAPSED_TIME = AIR_TIME+TAXI_IN+TAXI_OUT
* AIR_TIME = The time duration between wheels_off and wheels_on time
* DISTANCE = Distance between two airports
* CARRIER_DELAY = Delay caused by the airline in minutes
* WEATHER_DELAY = Delay caused by weather
* NAS_DELAY = Delay caused by air system
* SECURITY_DELAY = caused by security reasons
* LATE_AIRCRAFT_DELAY = Delay caused by security   

*Source : [kaggle](https://www.kaggle.com/datasets/yuanyuwendymu/airline-delay-and-cancellation-data-2009-2018?select=2009.csv)*

The primary objective of the project is to predict whether a flight is going to be delayed or not, **prior to this information being displayed on the departure board**. Therefore, to define delay, we will look at the column `ARR_DELAY`. If a flight arrives late at its destination, it will be considered as delayed. This is a numeric columns with the time difference in the arrival as minutes. Positive values indicate a delay, while a negative value indicates an early arrival. As first step, we will have a look at  the distribution of these levels in the dataset as that is an important aspect while training our models.
```{r}
nrow(subset(flight,ARR_DELAY>0))
nrow(subset(flight,ARR_DELAY<=0))
```

We can see that there are more flights that are not delayed than flights that are delayed. This might represent some imbalance in our dataset.

Let's create a new binary variable `DELAYED_STATUS` to indicate whether a flight was delayed `1` or not delayed `0`
```{r}
flight$DELAYED_STATUS <- if_else(flight$ARR_DELAY > 0, 1, 0)

table(flight$DELAYED_STATUS)
```

From the glossary, we can also identify a few attributes that give a lot of information on whether a flight is going to be delayed. These columns are:    

* Wheels On
* Taxi In
* Actual Elapsed Time   

It may not be a good idea to include these attributes during model building as this information is available after the flight takes off from the source airport and is not logical to make use of to predict whether the flight is going to be delayed. To demonstrate this idea, we can build two models - one without these attributes and one with them. The model with these attributes will result in better accuracy, precision, and recall.

“Departure Delay” is not listed above, which may be considered a borderline variable known just as the flight departs from source. However, departure delay does not always indicate a delay in arrival. It could be the case that a flight has departure delay but airlines try to account for these delays by reducing their elapse time, and in some cases, they succeed and the planes end up arriving on time or even earlier.    
We can see this with the graph below - 
```{r}
dummy1 <- flight %>% pivot_longer(.,cols = c('DEP_DELAY','ARR_DELAY'),names_to='value_category')
ggplot(dummy1,aes(fill=value_category,x=value,y=OP_CARRIER)) + geom_bar(position="dodge", stat="identity")+labs(title="Total Arrival/Departure Delay per Airline",x="Total Arrival/Departure Delay (mins)",y="Unique Airline Identifier")
```
We get a rough idea that departure delays do not necessarily amount to an arrival delay with this plot, as some airlines can be seen to have total departure delays greater than their arrival delays.   

Looking at the mean arrival/departure delays will reinforce the point further.
```{r}
group1 <- dummy1 %>% group_by(OP_CARRIER,value_category) %>% 
  summarize(
  mean_value = mean(value,na.rm=T)
)
ggplot(group1,aes(fill=value_category,x=mean_value,y=OP_CARRIER)) + geom_bar(position="dodge", stat="identity")+labs(title="Mean Arrival/Departure Delay per Airline",x="Mean Arrival/Departure Delay (mins)",y="Unique Airline Identifier")
```
We can clearly see that airlines that are prone to high departure delays, do not necessarily need to have a high arrival delay as well.

Before we proceed further, we noticed the presence of a column called `Unnamed..27`, and most of the rows at first glance are nulls. Looking for all the null values in this column.
```{r}
sum(is.na(flight$Unnamed..27))
```
We see that all rows for this column are null. Therefore, we can drop this column. 
```{r}
flight <- subset(flight,select=-c(Unnamed..27))
```

Before moving on, typing the column names in caps all the time is very exhausting. Therefore, we will further convert the column names to all lower cases.
```{r}
names(flight) <- tolower(names(flight))
xkabledplyhead(flight,title="The first 5 rows with lower case column names")
```

Aside from predicting arrival delay, as a secondary analysis we also want to predict flight cancellation prior to it being announced. We notice the presence of another feature in the dataset, that gives us this information - `cancelled`. However, flight_delay_status and cancellation cannot be used together at the same time. If a flight gets cancelled, then delay does not make sense and should not be applicable to our prediction. Let's make sure that this is the case. If a flight is marked as `cancelled` = 1, then arrival delay and departure delay should be null values.
```{r}
can_flight <- subset(flight,cancelled>0)
nrow(can_flight)
sum(is.na(can_flight$arr_delay))
```
From the above, we see that our assumption stands. For flights that were cancelled, arrival delay is always null. So to build a model that predicts whether a flight is delayed, we cannot use the data for when flights were cancelled as we will get erroneous results. Therefore, we will subset our dataset for only `cancelled` = 0.    
```{r}
flight <- subset(flight,cancelled==0)
flight <- subset(flight,select=-c(cancelled))
xkabledplyhead(flight,title="The first 5 rows of the new dataset")
```
In addition to this, there's another column which is called the `cancellation_code` which explains the cancellation reason for these cancelled flights. Since, we've already subset to `cancelled` = 0, we expect to see nulls in this column as there needn't be a cancellation reason any longer. Verifying this with below code.
```{r}
nrow(flight)
nrow(subset(flight,is.na(cancellation_code)||cancellation_code==""))
```
With the above code, we see that all values in `cancellation_code` are now null or blank values. Therefore, we can remove this column from the dataset.
```{r}
flight = subset(flight,select=-c(cancellation_code))
xkabledplyhead(flight,title="The first 5 rows of the dataset after removing cancellation_code")
```    

Another variable that catches attention is `diverted`. A diverted flight is one that has landed in a different airport than the one scheduled. This is normally temporary and the airline has the responsibility of taking passengers to their final destination without any extra cost. Therefore this is ideally considered a delayed flight.   
```{r}
unique(flight$diverted)
nrow(subset(flight,diverted==1))
nrow(subset(flight,diverted==0))
```
This column contains only 2 values, 0 and 1. The 1's are extremely small in occurrence.
The assumption here is that 1 indicates that a flight has been diverted. A diverted flight should have the `arr_delay` column populated, as it should definitely have delay on arrival to the actual destination airport. Confirming this with the below code.
```{r}
unique(subset(flight,diverted==1)$arr_delay)
```
The above results are inconsistent with the assumption. As it turns out, all diverted flights have no information on the `arr_delay`. This renders this column useless to us for predicting delay. Therefore, we will be removing this column as well.
```{r}
flight <- subset(flight,select=-c(diverted))
```

Now turning our attention to the the columns that cite the delay reason. On first glance, majority of the observations contain null values which is not a good sign. Lets get a count of how many null values there are.
```{r}
total <- nrow(flight)
total
(nrow(subset(flight,is.na(carrier_delay)))/total)*100
(nrow(subset(flight,is.na(weather_delay)))/total)*100
(nrow(subset(flight,is.na(nas_delay)))/total)*100
(nrow(subset(flight,is.na(security_delay)))/total)*100
(nrow(subset(flight,is.na(late_aircraft_delay)))/total)*100
```
We see that around 81.5% of all our observations contain null values for these columns and for some reason, the information is missing from columns at all the same instances. The conclusion is that we cannot use this column for any useful predictions owing to the large missing values. The choice is to drop these columns too. However, before doing that, we'll make a copy of flight dataset, maybe we can perform interesting EDA using this column using available rows (finding out the top reason for delay).
```{r}
flight_copy <- flight
flight <- subset(flight, select=-c(carrier_delay,weather_delay,nas_delay,security_delay,late_aircraft_delay))
```

## Checking for further missing values in the dataset

```{r}
colSums(is.na(flight))
```
We see that quite a few columns still have null values. However, the proportion of these values is extremely small as compared to the size of the dataset. We will go ahead and remove these rows to ensure model is built correctly.
```{r}
flight <- subset(flight,!is.na(arr_delay)&!is.na(taxi_in))
colSums(is.na(flight))
```

Checking the distribution of `delayed_status` column again.
```{r}
table(flight$delayed_status)
```
We still have enough 0's and 1's to build a logistic regression model.   

Now, for EDA purposes, we will make use of the external files airlines and airports to get respective airline names and city/state names into the dataset.

First getting the airline names.
```{r}
flight$airline <- airlines$AIRLINE[match(flight$op_carrier, airlines$IATA_CODE)]
```   

Next, getting the city name and state name into the dataset.
```{r}
flight$city_orig <- airports$CITY[match(flight$origin, airports$IATA_CODE)]
flight$city_dest <- airports$CITY[match(flight$dest, airports$IATA_CODE)]
flight$state_orig <- airports$STATE[match(flight$origin, airports$IATA_CODE)]
flight$state_dest <- airports$STATE[match(flight$dest, airports$IATA_CODE)]
```
We can drop the original columns because it is redundant information.
```{r}
flight <- subset(flight,select=-c(op_carrier))
```   

Since there are 308 total cities and 322 airports (multiple airports in a city) and 50 States in the dataset, we will not include this in the model because we'll get those many factor levels which will be hard to interpret. There are included only for EDA purpose.

Now to answer one of the SMART questions, since there is slight skewness between the predicted variable, accuracy will not be enough as a criteria to judge our model. Precision and recall will play an important role as well. However, as the business problem is predicting whether a flight is delayed or not, it is more important to catch all flights that are actually delayed as opposed to optimizing whether a predicted delay is correct or not. There is no harm in proactively anticipating and preparing for flights that have anticipated delayed and they turn out to be not delayed. However, there is greater business loss if we miss out on flights that delayed but we were not able to identify them.

In technical terms, our models need to be tuned to have a higher **recall** as opposed to **precision**.


## Feature Engineering and EDA   

We can also make a month and a day variable, which could be good predictors of delays. Also creating a day of the week variable starting with Sunday.
```{r}
flight$MONTH <- substr(flight$fl_date, 6, 7)
table(flight$MONTH)
flight$DAY <- substr(flight$fl_date, 9, 10)
#table(flight$DAY)
flight$dow <- lubridate::wday(flight$fl_date,label=TRUE,abbr=FALSE)
table(flight$dow)
```

We can also extract the hour of the scheduled flight to see if time of day is a good predictor
```{r}
# flight <- flight %>%
#   mutate(
#     hour = case_when(
#       nchar(crs_dep_time) == 3 ~ substr(crs_dep_time, 2, 3),
#       nchar(crs_dep_time) == 4 ~ substr(crs_dep_time, 3, 4),
#       TRUE ~ crs_dep_time
#     )
#   )

flight$crs_dep_time <- if_else(nchar(flight$crs_dep_time) != 4, paste0("0", flight$crs_dep_time), as.character(flight$crs_dep_time)) 
flight$crs_dep_time <- if_else(nchar(flight$crs_dep_time) != 4, paste0("0", flight$crs_dep_time), as.character(flight$crs_dep_time)) 

flight$hour <- as.integer(format(strptime(flight$crs_dep_time, "%H%M"), "%H"))

table(flight$hour)

flight %>% filter(hour == 00)
```

Lets first answer some questions on airline quality in terms of delays through some EDA.

Graph of Number of Flights by Airline
```{r}
flight %>%
  group_by(airline) %>%
  summarize(
    Count = n()
  ) %>%
  ggplot(
    aes(
      y = fct_reorder(airline, Count, .desc = T),
      x = Count
    )
  ) +
  geom_col(fill = "lightblue", color = "black") +
  labs(
    title = "Number of Flights by Airline",
    y = "Airline",
    x = "Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```
This is quite an interesting plot to start with as it already tells us which are the airlines that have the most domestic flights in the US:   

1. Southwest Airlines
2. Delta Airlines
3. American Airlines
4. SkyWest Airlines
5. Atlantic South East Airlines

Now the question to ask ourselves would be: if the top 5 airlines in terms of number of flight, are also the best in terms of arriving on time? We'll see which are the airlines with the most delayed flight to try to answer this:   

Graphing Number of Flight Delays by Airlines
```{r}
subset(flight,delayed_status==1) %>%
  group_by(airline) %>%
  summarize(
    Count = n()
  ) %>%
  ggplot(
    aes(
      y = fct_reorder(airline, Count, .desc = T),
      x = Count
    )
  ) +
  geom_col(fill = "lightblue", color = "black") +
  labs(
    title = "Number of Delayed Flights by Airline",
    y = "Airline",
    x = "Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```
From this plot we can now extract the top 5 airlines with the most delayed flights, which are:

1. Southwest Airlines
2. Delta Airlines
3. American Airlines
4. SkyWest Airlines
5. Atlantic SouthEast Airlines

It appears as if the Airlines with the most number of flights, tend to have more number of delayed flights as well.

However, this may not be the best way to judge the performance of an airlines. To do so, we will calculate the percentage of delayed flights and the average delay time per airline, as that might be a better representation of how the airlines really perform.

The first thing to calculate is the overall percentage of delayed flights so that we can do proper comparisons with each airline. In other words, airlines would want to be below the average percentage of delayed flights to be in an acceptable position, so this mean number will represent our threshold. Airlines above would be by common sense the ones that travelers would want to avoid as it means you those will have the most delays.

```{r}
threshold <- mean(flight$delayed_status)
round(threshold*100,digits=2)
```
This result goes to say that 36.51% of the flights are delayed, roughly 1 out of every 3 flight is delayed.

```{r}
test_1 <- flight %>% group_by(airline) %>% summarize(delays = sum(delayed_status))
test_2 <- flight %>% group_by(airline) %>% summarize(count = n())
dummy2 <- full_join(test_1,test_2,by='airline')
dummy2$percent_delay <- round((dummy2$delays/dummy2$count)*100,digits=2)
```   

Using this extra manipulated data, we can plot the percentage of flight delays per airline, also showcasing the threshold value that is calculated from the data. This should give us a good indicator of which airline performs well.  

```{r}
ggplot(data=dummy2, 
       aes(x=percent_delay,y=fct_reorder(airline, percent_delay, .desc = T)))+geom_col(fill = "lightblue", color = "black")+geom_vline(xintercept=36.51,color='blue',linetype='dashed',size=0.5)+labs(title="Percentage of delay per airline",x="Percent",y="Airline")
```

This is provides us with really interesting observations. 

Delta Airlines is the second airline with the highest number of flights, at the same time is again the second highest number of delayed flights, but in terms of percentage of delayed flights, it is the airline with the best percentage at 29.02%. That represents ~7% below the threshold.

American Airlines in another airline that tops the list in terms of number of flights but is still below the threshold value in terms of delay. So these two airlines must be the most sought after by passengers if they want to avoid delay.

Southwest Airlines has the most flights and consequently, a higher number of delayed flights as well, but as our last plot suggests, it is not the worst, it is the sixth best and is above the threshold value by over 1%, so definitely not great compared to Delta Airlines.   
Similarly Atlantic SouthEast Airlines and United airlines have large number of flights but perform somewhat moderately when compared to other airlines.   

Now coming to the bad airlines. The top 4 airlines with the highest percent of delays happen to be the top 4 airlines with the least number of flights as well. These airlines should definitely be avoided by passengers if they want to ensure timely arrival. 

Further airline testing. Plotting total number of delay in minutes by each airline
```{r}
flight %>% group_by(airline) %>% summarize(total = sum(arr_delay)) %>% ggplot(aes(x=total,fct_reorder(airline,total,.desc=T)))+geom_col(color='light blue',fill="pink") + labs(title="Total minutes of flight delay",x="Sum of minutes",y="Airline")
```
With the help of this plot, we see that Southwest Airlines has the highest flight delays in number of minutes. So a passenger flying this airline can expect to extremely high delays (in minutes) in their travel.   
Delta airline seems to do well here as well, being the 3rd highest in total delay duration, continuing to catch our attention!   

Another airline that comes into notice with this plot is the Alaska Airlines. The negative bar plot indicates that they are prone to actually arriving earlier than usual by a larger margin than their flights get delayed! Which is good news for the passengers, most of the times.   

Verdict: Airlines to travel by - Delta Airlines, Alaska Airlines.

Worst and best months to travel based on Flight delays.
```{r}
subset(flight,delayed_status==1) %>%
  group_by(MONTH) %>% summarize(count = n()) %>% ggplot(
    aes(
      x = fct_reorder(MONTH,count,.desc=T),
      y = count
    )
  ) +
  geom_col(fill = "lightblue", col = "black") +
  labs(
    title = "Number of Delayed Flights by Month",
    x = "Month",
    y = "Count"
  ) +
  theme_bw()
```
Seems like the best month to travel is September with the least number of Delays, whereas June consists of a lot of delays.

However, this is just an absolute number. A better way to find the best and worst months to travel will be through proportions. Lets first plot the count of flights that run in each month.
```{r}
flight %>%
  group_by(MONTH) %>% summarize(count = n()) %>% ggplot(
    aes(
      x = fct_reorder(MONTH,count,.desc=T),
      y = count
    )
  ) +
  geom_col(fill = "lightblue", col = "black") +
  labs(
    title = "Number of Flights by Month",
    x = "Month",
    y = "Count"
  ) +
  theme_bw()
```
Looks the busiest months are July, August and May. One possible explanation is start and end of summer vacations??

Now lets plot the proportion.
Graph of percent of flight delays by month
```{r}
flight %>%
  group_by(MONTH) %>%
  summarize(
    prop_delayed = mean(delayed_status, na.rm = T) * 100
  ) %>%
  ggplot(
    aes(
      x = fct_reorder(MONTH, prop_delayed, .desc = T),
      y = prop_delayed
    )
  ) +
  geom_col(fill = "lightblue", col = "black") +
  labs(
    title = "Percent of Flights Delayed by Month",
    x = "Month",
    y = "Percent"
  ) +
  theme_bw()
```
Seems like September runs the least number of delayed flights being the 9th busiest month and has the lowest proportion of delay. October, too is the 7th busiest month and 2nd least delay month. Seems like airline companies really buckle up on delays in these months whereas maintaining a moderately high number of flights.

January runs the 2nd least number of flights with the 4th highest delays. Must be stressful for the few passengers that travel then.

March and July run a large number of flights with a large number of delays.

Lets repeat the same exercise for day of the week. Graphing Number of Flight Delays by Day Of Week
```{r}
subset(flight,delayed_status==1) %>% group_by(dow) %>% summarize(count = n()) %>%
  ggplot(
    aes(
      y = fct_reorder(dow,count,.desc=T),
      x = count
    )
  ) +
  geom_col(fill = "lightblue", col = "black") +
  labs(
    title = "Number of Delayed Flights by DOW",
    y = "DOW",
    x = "Count"
  ) +
  theme_bw()
```
Now plotting the total number of flights in the week.
```{r}
flight %>%
  group_by(dow) %>% summarize(count = n()) %>% ggplot(
    aes(
      y = fct_reorder(dow,count,.desc=T),
      x = count
    )
  ) +
  geom_col(fill = "lightblue", col = "black") +
  labs(
    title = "Number of Flights by Day of Week",
    y = "DOW",
    x = "Count"
  ) +
  theme_bw()
```
Seems like the busiest days are Friday and Thursday and a result, they also have the most delayed number of flights. Whereas, Saturday and Sunday are least busiest with lesser number of delays.

Graph of percent of flight delays by Day of Week
```{r}
flight %>%
  group_by(dow) %>%
  summarize(
    prop_delayed = mean(delayed_status, na.rm = T) * 100
  ) %>%
  ggplot(
    aes(
      y = fct_reorder(dow, prop_delayed, .desc = T),
      x = prop_delayed
    )
  ) +
  geom_col(fill = "lightblue", col = "black") +
  labs(
    title = "Percent of Flights Delayed by DOW",
    y = "DOW",
    x = "Percent"
  ) +
  theme_bw()
```

Now lets try to see the top 20 most popular cities and how their average arrival delay looks like.   

Lets look at the 20 most popular destination cities in the US first.
```{r}
dummy4 <- flight %>% group_by(city_dest) %>% summarize(NOF = n(),Average_delay = mean(arr_delay)) 
dummy4 <- dummy4[order(dummy4$NOF,decreasing=TRUE),]
dummy4 <- dummy4[1:20,]

ggplot(dummy4,aes(x=NOF,y=fct_reorder(city_dest,NOF,.desc=T)))+geom_col(fill="pink",color="black")+labs(title="Top 20 most visited cities",x="Number of Flights",y="City")
```

Now lets add another axis and look at the average delay in the top 20 destinations.
```{r}
dummy6 <- dummy4 %>% pivot_longer(.,cols = c('NOF','Average_delay'),names_to='value_category') %>% mutate(scaled_value = ifelse(value_category=="Average_delay",value,value/10000))
ggplot(dummy6,aes(x=scaled_value,y=reorder(city_dest,value,.desc=T),fill=value_category)) + 
  geom_col(position="dodge") + 
  scale_x_continuous(sec.axis = sec_axis(~ . *10000 , name = "Number of Flights"))+
  labs(x="Average Arr Delay",y="City") 
```
This graph gives us lots of useful information about the top 20 most visited cities.
It is interesting to see how Atlanta, having such a high number of landings, has a very low average delay with just over 2 minutes, whereas Newark, a not so popular destination, has such a high minute average delay. 
New York, which is the 7th most popular destination (not sure why this is the case since its so popular) has the highest average delay out of all, which doesn't add up with it being the 7th most visited place.     San Francisco is another destination that stands out with a high average delay as well as Orlando and Boston.   

Fun Fact - Salt Lake City is famous for its early arriving flight. We should probably go and visit.   

Now lets repeat the same time for top 20 origin cities and associated departure delays with them.
```{r}
dummy5 <- flight %>% group_by(city_orig) %>% summarize(NOF = n(),Average_delay = mean(dep_delay)) 
dummy5 <- dummy5[order(dummy5$NOF,decreasing=TRUE),]
dummy5 <- dummy5[1:20,]
dummy7 <- dummy5 %>% pivot_longer(.,cols = c('NOF','Average_delay'),names_to='value_category') %>% mutate(scaled_value = ifelse(value_category=="Average_delay",value,value/10000))
ggplot(dummy7,aes(x=scaled_value,y=reorder(city_orig,value,.desc=T),fill=value_category)) + 
  geom_col(position="dodge") + 
  scale_x_continuous(sec.axis = sec_axis(~ . *10000 , name = "Number of Flights"))+
  labs(x="Average Dep Delay",y="City") 

```
This is an important graph when compared to the previous one. It seems like majorly, a city have a high departure delay, corresponds to a high arrival delay as we can see with new york, newark etc. This may lead someone to believe that departure delay always corresponds to arrival delay.    
However, Salt Lake City to the rescue to save us from this erroneous assumption. Despite having moderately high arrival delay, the flights to this city are somehow able to compensate for this and arrive earlier on average. 

For the last EDA, lets also plot the number of destinations by Airlines to see which airline has the maximum 'coverage'.
```{r}
flight %>% group_by(airline) %>% summarize(dest = length(unique(city_dest))) %>% ggplot(aes(x=dest,y=fct_reorder(airline,dest,.desc=T)))+geom_col(fill="pink", color="black")+labs(title="Number of Destinations per Airline",x="Count of Destinations",y="Airline")
```
Delta Air Lines has once again managed to appear in top 3, providing with the 3rd highest coverage. Its a no brainer at this point to use this airline for domestic travel in the US.

# Modeling

## Logistic Regression

Having over 500,000 is too much to process in the model and is causing the session to slow down. To prevent this, we are going to further subset the dataset before processing.
```{r}
set.seed(333)
sample_size = round(nrow(flight)*0.05)
flight_small <- flight[sample(nrow(flight), sample_size, replace = F),]
head(flight_small)
```


```{r}
flight_small$airline <- as.factor(flight_small$airline)
flight_small$city_orig <- as.factor(flight_small$city_orig)
flight_small$city_dest <- as.factor(flight_small$city_dest)
flight_small$month <- as.factor(flight_small$MONTH)

# flight_small <- flight_small[!is.na(flight_small$op_carrier),]
# flight_small <- flight_small[!is.na(flight_small$origin),]
# flight_small <- flight_small[!is.na(flight_small$dest),]
# flight_small <- flight_small[!is.na(flight_small$month),]
# flight_small <- flight_small[!is.na(flight_small$delayed_status),]

str(flight_small)
```

We will include the following variables in our initial development of our model to predict delayed status:
* Airline
* Origin Airport
* Destination Airport
* Month
* Day of Week
* Time of Day (by hour of departure)
* Distance

Train-test Split of Data
```{r}
set.seed(322)
flight_sample <- sample(2, nrow(flight_small), replace = TRUE, prob = c(0.8, 0.2))

flight_training <- flight_small[flight_sample == 1, ]
flight_test <- flight_small[flight_sample == 2, ]
```


```{r}
logit <- glm(delayed_status ~ airline + origin + dest + MONTH + dow + hour + distance, data = flight_training, family = "binomial")
summary(logit)
```

The following coefficients are significant:
```{r}
k <- summary(logit)$coefficients[,4]
subset(k, k < 0.05)
```

*ADD THE REST*
The following Airlines are significant:
* Frontier Airlines
* Skywest Airlines
* Southwest Airlines
* Spirit Airlines

The following months are significant:
* February
* June
* October
* December

The following days of the week:
* Q?
* 6?

Hour was also a significant predictor.

Interestingly, none of the origin or destination airports were significant predictors of delays.

Let's plot the ROC
```{r}
loadPkg("pROC") 
prob <- predict(logit, type = "response" )
flight_training$prob <- prob
h <- roc(delayed_status ~ prob, data = flight_training)
auc(h) # 0.654
plot(h)
```

```{r}
flight_test <- flight_test[!(flight_test$origin == "SCE" | flight_test$origin == "TXK"), ]
flight_test <- flight_test[!(flight_test$dest == "GRI" | flight_test$dest == "WYS"), ]

loadPkg("pROC") 
prob <- predict(logit, flight_test, type = "response" )
flight_test$prob <- prob
h <- roc(delayed_status ~ prob, data = flight_test)
auc(h) # 0.60
plot(h)
```

Area under the curve is 0.6543 on the training data, and 0.60 on the testing data

```{r}
library(pscl)
pR2(logit)
```

McFadden value is 0.058

Confusion Matrix
```{r}
library(caret)

flight_test$response <- as.factor(if_else(flight_test$prob >=0.5, 1, 0))
flight_test$delayed_status <- as.factor(flight_test$delayed_status)

confusionMatrix(flight_test$response, flight_test$delayed_status)
```

Accuracy on model: 62.9%
Sensitivity / Recall: 19.1% - 409 / (1731+409)
Specificity: 88.57%
Precision: 49.5%

The specificity on this model is pretty good, but the other metric could use some improvment. Let's try with a different cutoff value
```{r}
flight_test$response <- as.factor(if_else(flight_test$prob >=0.4, 1, 0))
flight_test$delayed_status <- as.factor(flight_test$delayed_status)

confusionMatrix(flight_test$response, flight_test$delayed_status)
```

At a lower cutoff value, the accuracy is slightly lower at 59.8% but there is an improvement in the sensitivity of the model at 45.5%

### Trying with other variables
```{r}
logit2 <- glm(delayed_status ~ airline + origin + dest + MONTH + dow + hour + distance + taxi_in + wheels_on + actual_elapsed_time, data = flight_training, family = "binomial")
summary(logit2)
```

Significant Coefficients
```{r}
k <- summary(logit)$coefficients[,4]
subset(k, k < 0.05)
```

McFadden
```{r}
pR2(logit)
```

Plotting ROC
```{r}
prob <- predict(logit, type = "response" )
flight_training$prob <- prob
h <- roc(delayed_status ~ prob, data = flight_training)
auc(h) # 0.6543
plot(h)
```

Confusion Matrix
```{r}
flight_test$response <- as.factor(if_else(flight_test$prob >=0.5, 1, 0))
flight_test$delayed_status <- as.factor(flight_test$delayed_status)

confusionMatrix(flight_test$response, flight_test$delayed_status)
```


## Random Forest
Train-test splitting

Unable to use factor variables that are more than 53 categories, so we have to drop origin and destination airports
```{r}
set.seed(333)
sample_size = round(nrow(flight)*0.01)
flight_small <- flight[sample(nrow(flight), sample_size, replace = F),]

set.seed(322)
flight_sample <- sample(2, nrow(flight_small), replace = TRUE, prob = c(0.8, 0.2))

flight_training <- flight_small[flight_sample == 1, ]
flight_test <- flight_small[flight_sample == 2, ]

X_train <- subset(flight_training, select = c(airline, MONTH, dow, hour, distance, delayed_status))
X_test <-subset(flight_test, select = c(airline, MONTH, dow, hour, distance, delayed_status))

X_train$delayed_status <- as.factor(X_train$delayed_status)
X_train$MONTH <- as.factor(X_train$MONTH)
X_test$delayed_status <- as.factor(X_test$delayed_status)
X_test$MONTH <- as.factor(X_test$MONTH)

str(X_train)
```

Building Model
```{r}
library(randomForest)

set.seed(64)
rf <- randomForest(delayed_status ~ ., data = X_train, proximity = TRUE)
rf
```

Making Confusion Matricies
```{r}
p1 <- predict(rf, X_train)
confusionMatrix(p1, X_train$delayed_status)
```                

99.9% Accuracy on the training set

Confusion Matrix for the Testing set
```{r}
p2 <- predict(rf, X_test)
confusionMatrix(p2, X_test$delayed_status)
```

Accuracy: 57.8%
Sensitivity / Recall: 27.4%
Specificity: 75.0%
Precision: 38.2%

Again, specificity is good but the other measures are not.

Trying with different amounts of trees
```{r}
set.seed(99)
rf2 <- randomForest(delayed_status ~ ., data = X_train, proximity = TRUE, maxnodes = 10)

p3 <- predict(rf, X_test)
confusionMatrix(p3, X_test$delayed_status)
```


