---
title: "Airline Delay and Cancellation"
author: "Group 5"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F,results='markup')

# importing libraries
library(ezids)
library(ggplot2)
library(tidyverse)
library(dplyr)
```

Importing data
```{r}
airlines <- data.frame(read.csv("data/airlines.csv"))
airports <- data.frame(read.csv("data/airports.csv"))
#flights <- data.frame(read.csv("C:/Users/Aditya Kumar/OneDrive/Study/GWU - Data Science/DATS 6101 Intro to Data Science/Project 02/archive (1)/2015.csv"))
#nrow(flights)
```

Our entire flights data for 2015 has 5,819,079 rows which is extremely large. For the sake of our analysis, we will only randomly select 581,908 rows which is 10% of the original dataset size.    
Sub-setting dataset because it is very large. 
```{r}
#set.seed(333)
#sample_size = round(nrow(flights)*0.1)
#flights_sample <- flights[sample(nrow(flights), sample_size, replace = F),]
#head(flights_sample)
```

Exporting smaller sample
```{r}
#write.csv(flights_sample, file = "data/flights_sample.csv",row.names=F)

flight <- read.csv("data/flights_sample.csv")
```

```{r}
xkabledplyhead(flight,title="The first 5 rows of the dataset")
str(flight)
```

This data now has `r nrow(flight)` rows.   

Now we'll begin our analysis.   

# Pre-processing and initial analysis   

Glossary of all the columns in the dataset:   

* FL_DATE = Date of the Flight
* OP_CARRIER = Airline Identifier
* OP_CARRIER_FL_NUM = Flight Number
* ORIGIN = Starting Airport Code
* DEST = Destination Airport Code
* CRS_DEP_TIME = Planned Departure Time
* DEP_TIME = Actual Departure Time
* DEP_DELAY = Total Delay on Departure in minutes
* TAXI_OUT = The time duration elapsed between departure from the origin airport gate and wheels off
* WHEELS_OFF = The time point that the aircraft's wheels leave the ground
* WHEELS_ON = The time point that the aircraft's wheels touch on the ground
* TAXI_IN = The time duration elapsed between wheels-on and gate arrival at the destination airport
* CRS_ARR_TIME = Planned arrival time
* ARR_TIME = Actual Arrival Time = ARRIVAL_TIME - SCHEDULED_ARRIVAL
* ARR_DELAY = Total Delay on Arrival in minutes
* CANCELLED = Flight Cancelled (1 = cancelled)
* CANCELLATION_CODE = Reason for Cancellation of flight: A - Airline/Carrier; B - Weather; C - National Air System; D - Security
* DIVERTED = Aircraft landed on different airport that the one scheduled
* CRS_ELAPSED_TIME = Planned time amount needed for the flight trip
* ACTUAL_ELAPSED_TIME = AIR_TIME+TAXI_IN+TAXI_OUT
* AIR_TIME = The time duration between wheels_off and wheels_on time
* DISTANCE = Distance between two airports
* CARRIER_DELAY = Delay caused by the airline in minutes
* WEATHER_DELAY = Delay caused by weather
* NAS_DELAY = Delay caused by air system
* SECURITY_DELAY = caused by security reasons
* LATE_AIRCRAFT_DELAY = Delay caused by security   

*Source : [kaggle](https://www.kaggle.com/datasets/yuanyuwendymu/airline-delay-and-cancellation-data-2009-2018?select=2009.csv)*

The primary objective of the project is to determine whether a flight is going to be delayed or not, **prior to this information being displayed on the departure board**. Therefore, to define delay, we will look at the column `ARR_DELAY`. If a flight arrives late at its destination, it will be considered as delayed. This is a numeric columns with the time difference in the arrival as minutes. Positive values indicate a delay, while a negative value indicates an early arrival. As first step, we'll have a look at  the distribution of these levels in the dataset as that is an important aspect while training our models.
```{r}
nrow(subset(flight,ARR_DELAY>0))
nrow(subset(flight,ARR_DELAY<=0))
```

We can see that there are more flights that are not delayed than flights that are delayed. This might represent some imbalance in our dataset.

Let's create a new binary variable `DELAYED_STATUS` to indicate whether a flight was delayed `1` or not delayed `0`
```{r}
flight$DELAYED_STATUS <- if_else(flight$ARR_DELAY > 0, 1, 0)

table(flight$DELAYED_STATUS)
```

From the glossary, we can also identify a few attributes that give a lot of information on whether a flight is going to be delayed. These columns are:    

* Wheels On
* Taxi In
* Actual Elapsed Time   

It may not be a good idea to include these attributes during model building as this information is available after the flight takes off from the source airport and is not logical to make use of to predict whether the flight is going to be delayed. To demonstrate this idea, we can build two models - one without these attributes and one with them. The model with these attributes will result in better accuracy, precision and recall.

“Departure Delay” is not listed above, which may be considered a borderline variable known just as the flight departs from source. However, departure delay does not always indicate a delay in arrival. It could be the case that a flight has departure delay but airlines try to account for these delays by reducing their elapse time, and in some cases, they succeed and the planes end up arriving on time or even earlier.    
We can see this with the graph below - 
```{r}
dummy1 <- flight %>% pivot_longer(.,cols = c('DEP_DELAY','ARR_DELAY'),names_to='value_category')
ggplot(dummy1,aes(fill=value_category,x=value,y=OP_CARRIER)) + geom_bar(position="dodge", stat="identity")+labs(title="Total Arrival/Departure Delay per Airline",x="Total Arrival/Departure Delay (mins)",y="Unique Airline Identifier")
```
We get a rough idea that departure delays do not necessarily amount to an arrival delay with this plot, as some airlines can be seen to have total departure delays greater than their arrival delays.   

Looking at the mean arrival/departure delays will reinforce the point further.
```{r}
group1 <- dummy1 %>% group_by(OP_CARRIER,value_category) %>% 
  summarize(
  mean_value = mean(value,na.rm=T)
)
ggplot(group1,aes(fill=value_category,x=mean_value,y=OP_CARRIER)) + geom_bar(position="dodge", stat="identity")+labs(title="Mean Arrival/Departure Delay per Airline",x="Mean Arrival/Departure Delay (mins)",y="Unique Airline Identifier")
```
We can clearly see that airlines that are prone to high departure delays, do not necessarily need to have a high arrival delay as well.

Before we proceed further, we noticed the presence of a column called `Unnamed..27`, and most of the rows at first glance are nulls. Looking for all the null values in this column.
```{r}
sum(is.na(flight$Unnamed..27))
```
We see that all rows for this column are null. Therefore, we can drop this column. 
```{r}
flight <- subset(flight,select=-c(Unnamed..27))
```

Before moving on, typing the column names in caps all the time is very exhausting. Therefore, we will further convert the column names to all lower cases.
```{r}
names(flight) <- tolower(names(flight))
xkabledplyhead(flight,title="The first 5 rows with lower case column names")
```

Aside from predicting arrival delay, as a secondary analysis we also want to predict flight cancellation prior to it being announced. We notice the presence of another feature in the dataset, that gives us this information - `cancelled`. However, flight_delay_status and cancellation cannot be used together at the same time. If a flight gets cancelled, then delay does not make sense and shouldn't be applicable. Lets make sure that this is the case. If a flight is marked as `cancelled` = 1, then arrival delay and departure delay should be null values.
```{r}
can_flight <- subset(flight,cancelled>0)
nrow(can_flight)
sum(is.na(can_flight$arr_delay))
```
From the above, we see that our assumption stands. For flights that were cancelled, arrival delay is always null. So to build a model that predicts whether a flight is delayed, we cannot use the data for when flights were cancelled as we will get erroneous results. Therefore, we will subset our dataset for only `cancelled` = 0.    
```{r}
flight <- subset(flight,cancelled==0)
flight <- subset(flight,select=-c(cancelled))
xkabledplyhead(flight,title="The first 5 rows of the new dataset")
```
In addition to this, there's another column which is called the `cancellation_code` which explains the cancellation reason for these cancelled flights. Since, we've already subset to `cancelled` = 0, we expect to see nulls in this column as there needn't be a cancellation reason any longer. Verifying this with below code.
```{r}
nrow(flight)
nrow(subset(flight,is.na(cancellation_code)||cancellation_code==""))
```
With the above code, we see that all values in `cancellation_code` are now null or blank values. Therefore, we can remove this column from the dataset.
```{r}
flight = subset(flight,select=-c(cancellation_code))
xkabledplyhead(flight,title="The first 5 rows of the dataset after removing cancellation_code")
```    

Another variable that catches attention is `diverted`. A diverted flight is one that has landed in a different airport than the one scheduled. This is normally temporary and the airline has the responsibility of taking passengers to their final destination without any extra cost. Therefore this is ideally considered a delayed flight.   
```{r}
unique(flight$diverted)
nrow(subset(flight,diverted==1))
nrow(subset(flight,diverted==0))
```
This column contains only 2 values, 0 and 1. The 1's are extremely small in occurrence.
The assumption here is that 1 indicates that a flight has been diverted. A diverted flight should have the `arr_delay` column populated, as it should definitely have delay on arrival to the actual destination airport. Confirming this with the below code.
```{r}
unique(subset(flight,diverted==1)$arr_delay)
```
The above results are inconsistent with the assumption. As it turns out, all diverted flights have no information on the `arr_delay`. This renders this column useless to us for predicting delay. Therefore, we will be removing this column as well.
```{r}
flight <- subset(flight,select=-c(diverted))
```

Now turning our attention to the the columns that cite the delay reason. On first glance, majority of the observations contain null values which is not a good sign. Lets get a count of how many null values there are.
```{r}
total <- nrow(flight)
total
(nrow(subset(flight,is.na(carrier_delay)))/total)*100
(nrow(subset(flight,is.na(weather_delay)))/total)*100
(nrow(subset(flight,is.na(nas_delay)))/total)*100
(nrow(subset(flight,is.na(security_delay)))/total)*100
(nrow(subset(flight,is.na(late_aircraft_delay)))/total)*100
```
We see that around 81.5% of all our observations contain null values for these columns and for some reason, the information is missing from columns at all the same instances. The conclusion is that we cannot use this column for any useful predictions owing to the large missing values. The choice is to drop these columns too. However, before doing that, we'll make a copy of flight dataset, maybe we can perform interesting EDA using this column using available rows (finding out the top reason for delay).
```{r}
flight_copy <- flight
flight <- subset(flight, select=-c(carrier_delay,weather_delay,nas_delay,security_delay,late_aircraft_delay))
```

## Checking for further missing values in the dataset

```{r}
colSums(is.na(flight))
```
We see that quite a few columns still have null values. However, the proportion of these values is extremely small as compared to the size of the dataset. We will go ahead and remove these rows to ensure model is built correctly.
```{r}
flight <- subset(flight,!is.na(arr_delay)|!is.na(taxi_in))
colSums(is.na(flight))
```

Checking the distribution of `delayed_status` column again.
```{r}

```

## EDA   

We can also make a month and a day variable, which could be good predictors of delays.
```{r}
flight$MONTH <- substr(flight$fl_date, 6, 7)
table(flight$MONTH)
flight$DAY <- substr(flight$fl_date, 9, 10)
```

Graphing Number of Flight Delays by Month
```{r}
subset(flight,delayed_status==1) %>%
  ggplot(
    aes(
      x = MONTH
    )
  ) +
  geom_bar(fill = "lightblue", col = "black") +
  labs(
    title = "Number of Delayed Flights by Month",
    x = "Month",
    y = "Count"
  ) +
  theme_bw()
```

Graph of percent of flight delays by month
```{r}
flight %>%
  group_by(MONTH) %>%
  summarize(
    prop_delayed = mean(delayed_status, na.rm = T) * 100
  ) %>%
  ggplot(
    aes(
      x = MONTH,
      y = prop_delayed
    )
  ) +
  geom_col(fill = "lightblue", col = "black") +
  labs(
    title = "Percent of Flights Delayed by Month",
    x = "Month",
    y = "Percent"
  ) +
  theme_bw()
```

Graph of number of flights by carrier
```{r}
flight %>%
  group_by(op_carrier) %>%
  summarize(
    Count = n()
  ) %>%
  ggplot(
    aes(
      x = fct_reorder(op_carrier, Count, .desc = T),
      y = Count
    )
  ) +
  geom_col(fill = "lightblue", color = "black") +
  labs(
    title = "Number of Flights by Airline",
    x = "Airline",
    y = "Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

Percentage of Flight delays
```{r}
flight %>%
  group_by(op_carrier) %>%
  summarize(
    prop_delayed = mean(delayed_status, na.rm = T) * 100
  ) %>%
  ggplot(
    aes(
      x = fct_reorder(op_carrier, prop_delayed, .desc = T),
      y = prop_delayed
    )
  ) + 
  geom_col(fill = "lightblue", color = "black") + 
  labs(
    title = "Percent of Flights Delayed by Airline",
    x = "Airline",
    y = "Percent Delayed"
  ) + 
  theme_bw()
```

# Modeling

## Logistic Regression

Having over 500,000 is too much to process in the model and is causing the session to slow down. I'm going to subset it even smaller
```{r}
set.seed(333)
sample_size = round(nrow(flight)*0.01)
flight_small <- flight[sample(nrow(flight), sample_size, replace = F),]
head(flight_small)
```


```{r}
flight_small$op_carrier <- as.factor(flight_small$op_carrier)
flight_small$origin <- as.factor(flight_small$origin)
flight_small$dest <- as.factor(flight_small$dest)
flight_small$month <- as.factor(flight_small$MONTH)

flight_small <- flight_small[!is.na(flight_small$op_carrier),]
flight_small <- flight_small[!is.na(flight_small$origin),]
flight_small <- flight_small[!is.na(flight_small$dest),]
flight_small <- flight_small[!is.na(flight_small$month),]
flight_small <- flight_small[!is.na(flight_small$delayed_status),]

str(flight_small)
```


```{r}
logit <- glm(delayed_status ~ op_carrier + origin + dest + month, data = flight_small, family = "binomial")
summary(logit)
```

Plotting ROC
```{r}
loadPkg("pROC") 
prob <- predict(logit, type = "response" )
flight_small$prob <- prob
h <- roc(delayed_status ~ prob, data = flight_small)
auc(h) # 0.6944
plot(h)
```







