---
title: "Airline Delay and Cancellation"
author: "Group 5"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# importing libraries
library(ezids)
library(tidyverse)
```

Importing data
```{r}
airlines <- data.frame(read.csv("data/airlines.csv"))
airports <- data.frame(read.csv("data/airports.csv"))
#flights <- data.frame(read.csv("C:/Users/Aditya Kumar/OneDrive/Study/GWU - Data Science/DATS 6101 Intro to Data Science/Project 02/archive (1)/2015.csv"))
#nrow(flights)
```

Our entire flights data for 2015 has 5,819,079 rows which is extremely large. For the sake of our analysis, we will only randomly select `r round(nrow(flights)*0.1)` rows.    
Sub-setting dataset because it is very large. 
```{r}
#set.seed(333)
#sample_size = round(nrow(flights)*0.1)
#flights_sample <- flights[sample(nrow(flights), sample_size, replace = F),]
#head(flights_sample)
```

Exporting smaller sample
```{r}
#write.csv(flights_sample, file = "data/flights_sample.csv",row.names=F)

flight <- read.csv("data/flights_sample.csv")
```

```{r}
str(flight)
```


This data now has `r nrow(flight)` rows.   

Now we'll begin our analysis.   

# Pre-processing and initial analysis    

The primary objective of the project is to determine whether a flight is going to be delayed or not. Therefore, to define delay, we will look at the column `ARR_DELAY`. If a flight arrives late at its destination, it will be considered as delayed. This is a numeric columns with the time difference in the arrival as minutes. Positive values indicate a delay, while a negative value indicates an early arrival. As first step, we'll have a look at  the distribution of these levels in the dataset as that is an important aspect while training our models.
```{r}
nrow(subset(flight,ARR_DELAY>0))
nrow(subset(flight,ARR_DELAY<=0))
```

We can see that there are more flights that are not delayed than flights that are delayed. This might represent some imbalance in our dataset.

Let's create a new binary variable `DELAYED_STATUS` to indicate whether a flight was delayed `1` or not delayed `0`
```{r}
flight$DELAYED_STATUS <- if_else(flight$ARR_DELAY > 0, 1, 0)

table(flight$DELAYED_STATUS)
```

We can also make a month variable, which could be a good predictor of delays
```{r}
flight$MONTH <- substr(flight$FL_DATE, 6, 7)
table(flight$MONTH)
```

```{r}
subset(flight,DELAYED_STATUS==1) %>%
  ggplot(
    aes(
      x = MONTH
    )
  ) +
  geom_bar() +
  labs(
    title = "Number of Delayed Flights by Month",
    x = "Month",
    y = "Count"
  ) +
  theme_bw()
```

```{r}
flight %>%
  group_by(MONTH) %>%
  summarize(
    prop_delayed = mean(DELAYED_STATUS, na.rm = T) * 100
  ) %>%
  ggplot(
    aes(
      x = MONTH,
      y = prop_delayed
    )
  ) +
  geom_col() +
  labs(
    title = "Percent of Flights Delayed by Month",
    x = "Month",
    y = "Percent"
  ) +
  theme_bw()
```

