---
title: "Airline Delay and Cancellation"
author: "Group 5"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# importing libraries
library(ezids)
library(tidyverse)
```

Importing data
```{r}
airlines <- data.frame(read.csv("data/airlines.csv"))
airports <- data.frame(read.csv("data/airports.csv"))
#flights <- data.frame(read.csv("C:/Users/Aditya Kumar/OneDrive/Study/GWU - Data Science/DATS 6101 Intro to Data Science/Project 02/archive (1)/2015.csv"))
#nrow(flights)
```

Our entire flights data for 2015 has 5,819,079 rows which is extremely large. For the sake of our analysis, we will only randomly select `r round(nrow(flights)*0.1)` rows.    
Sub-setting dataset because it is very large. 
```{r}
#set.seed(333)
#sample_size = round(nrow(flights)*0.1)
#flights_sample <- flights[sample(nrow(flights), sample_size, replace = F),]
#head(flights_sample)
```

Exporting smaller sample
```{r}
#write.csv(flights_sample, file = "data/flights_sample.csv",row.names=F)

flight <- read.csv("data/flights_sample.csv")
```

This data now has `r nrow(flight)` rows.   

Now we'll begin our analysis.   

# Pre-processing and initial analysis    

The primary objective of the project is to determine whether a flight is going to be delayed or not. Therefore, to define delay, we will look at the column `ARR_DELAY`. If a flight arrives late at its destination, it will be considered as delayed. This is a factor column with two levels - 0 and 1. As first step, we'll have a look at  the distribution of these levels in the dataset as that is an important aspect while training our models.
```{r}
nrow(subset(flight,ARR_DELAY==1))
nrow(subset(flight,ARR_DELAY==0))
```
We can see that both levels of the dependent variable are equally distributed.   

